Transient dynamics of granular slopes: continuum v discrete modelling

### 2016-07-14 paper.tex
Normalised energy plots

\title{Transient dynamics of granular piles: continuum v discrete modelling}

@@

Granular flows have been studied in laboratory experiments in different 
geometries such as tilted piles leading to slope failure and surface
avalanches~\citep{Legros2002, Iverson1997} or by considering vertical columns 
of grains collapsing and spreading under their own
weight~\citep{Lajeunesse2004, Lajeunesse2005}. The dynamics observed in the
experiments is often nontrivial in the sense that the final configurations
after the dissipation of the whole kinetic energy can not be readily predicted
by means of simple laws based on the Mohr-Coulomb nature of the material. For
example, in collapsing columns, the runout distance is found to obey a
power-law dependence upon the initial aspect ratio of the column.

@@

In MPM simulations, the material point spacing is adopted to be the same as the 
mean grain diameter in DEM. A mesh size of 0.0125m is adopted with 25 material 
points per cell. The effect of mesh size and the number of material points per 
cell was investigated by~\citep{soundararajan2015}. The initial configuration
of the slope in MPM is shown in~\cref{fig:mpm_slope_setup}. Frictional boundary
conditions are applied on the left and the bottom boundaries by applying
constraints to the nodal acceleration. Mohr-Coulomb model with no dilation is
used to simulate the continuum behaviour of the granular pile. Periodic shear
tests in DEM reveals a macroscopic friction coefficient of 0.22.

@@

\Cref{fig:mpm_gradvelocity_200j} shows the initial evolution of the 
granular slope subjected to an initial horizontal energy $E_0 = 61$ (in 
dimensionless units) using MPM. As the granular slope is sheared along the 
bottom, the shear propagates to the top leaving a cavity in the vicinity of the 
left wall. This cavity gets partially filled as the granular mass at the top 
collapse behind the flowing mass due to inertia. Similar behaviour is observed 
during the initial stages of the flow evolution using CD technique 
(\cref{fig:Gradient_Slope_CD_200J}). Due to inertia, the grains at the top 
of the granular heap roll down to fill the cavity, while the pile continues to 
spread.

\begin{figure}[tbph]
  \centering
  \begin{subfigure}[b]{0.47\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figs/mpm_gradvelocity_200j}
    \caption{MPM simulation of the initial stages of granular pile subjected to a 
    gradient horizontal energy.}
    \label{fig:mpm_gradvelocity_200j}
  \end{subfigure}
  \begin{subfigure}[b]{0.47\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figs/Gradient_Slope_CD_200J}
    \caption{CD simulation of the initial stages of granular pile subjected to a 
    gradient horizontal energy.~\citep{Mutabaruka2013}.}
    \label{fig:Gradient_Slope_CD_200J}
  \end{subfigure}
  \label{fig:gradvelocity_200j}
\end{figure}

The flow involves a transient phase with a change in the geometry of the pile 
followed by continuous spreading. The gradient of input energy applied to the 
granular slope mimics a horizontal quake. Despite the creation 
of a cavity behind the flowing mass, the granular heap remains in 
contact with the left wall irrespective of the input 
energy.~\Cref{fig:runout_eo_mpm_dem} shows the normalised run-out distance 
$(L_f - L_0)/L_0$ and total run-out time $t_f$ as a function of the input 
energy $E_0$. Two regimes characterised by a power-law relation between 
the run-out distance and time as a function of $E_0$ can be observed. In the 
first regime, corresponding to the range of low input energies $E_0 < 40 \ 
mgd$, the run-out distance observed varies as $L_f \propto (E_0)^\alpha$ with 
$\alpha 
\simeq 0.206 \pm 0.012$ over nearly one decade. Overall, the run-out distance 
predicted by the continuum approach matches the DEM simulations. At very low 
energies, DEM simulations show longer run-out distance due to local 
fluidisation. The difference in the run-out between DEM and CD arise mainly 
from the scales of description and the inelastic nature of Contact Dynamics. 
Similar behaviour between DEM and CD approaches was observed 
by~\citet{Radjai1997}.

@@

In the second regime, corresponding to the range of high input energies  $E_0 > 
40 \ mgd$, the run-out distance varies as $L_f \propto (E_0)^{\alpha'}$ over 
one decade with $\alpha' \simeq 0.56\pm 0.04$ while the duration increases as 
$t_f \propto (E_0)^{\beta'}$ with $\beta' \simeq 0.33 \pm 0.02$. Hence, in this 
regime the average run-out speed varies as $v_s \propto (E_0)^{0.498 \pm 
0.01}$. This exponent is close to the value $0.5$ in $v_0 \propto (E_0)^{0.5}$, 
and hence, within the confidence interval of the exponents.
%, in the second 
%regime we may assume $\beta' \simeq \alpha' - 0.5$ and $v_s \propto v_0$. 
In the second regime, both DEM and MPM predict almost the same run-out 
behaviour. However, MPM predicts longer duration with increase in the input 
energy.

@@

The evolution of total kinetic energies $E_k$ with time for different values of 
the input energy $E_{ki}$ based on MPM simulations are shown 
in~\cref{fig:Energy_Time_Slope}. The MPM simulation shows two distinct regimes 
in the normalised kinetic energy plot as a function of normalised time 
in~\cref{fig:Normalised_Energy_Time_Slope}. However, the DEM simulations 
(\cref{fig:Normalised_Energy_Time_Slope_DEM}) show that the energy evolution 
corresponding to the low energy regime nearly collapse on to a single time 
evolution. This is consistent with the observation of run-out time $t_f$ being 
independent of the input energy. In contrast, MPM simulations predict a power 
law relation between the run-out duration and input energy. However, the 
plots corresponding to the high energy regime (\cref{fig:Energy_Time_Slope}), 
collapse only at the beginning of the run-out i.e. for $t < t_1 \simeq 7.5 \ 
(d/g)^{0.5}$. Although MPM simulations show a longer duration of run-out 
(\cref{fig:Energy_Time_Slope}), the total kinetic energy is completely 
dissipated at $t = 60 \sqrt{d/g}$. DEM simulations predict $t = 80 \sqrt{d/g}$ 
for the kinetic energy to be completely dissipated. This is due to grain 
rearrangement at the free surface (\cref{fig:voro_500}). The granular mass 
densifies as the flow progresses, after the initial dilation phase for $t = 
20\sqrt{d/g}$.

\begin{figure}[tbhp]
  \centering
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/Energy_Slope}
    \caption{Evolution of total kinetic energy with time.}
    \label{fig:energy_slope}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figs/Normalised_Energy_Time_Slope}
    \caption{Evolution of normalised kinetic energy with normalised time.}
    \label{fig:Normalised_Energy_Time_Slope}
  \end{subfigure}
  \caption{Evolution of kinetic energy with time (MPM) for a pile subjected to 
  gradient input velocities.}
  \label{fig:Energy_Time_Slope}
\end{figure}

\begin{figure}[tbhp]
  \centering
  \includegraphics[width=0.9\textwidth]{figs/Normalised_Energy_Time_Slope_DEM}
  \caption{Evolution of normalised kinetic energy with normalised time for a pile 
  subjected to gradient input velocities.}
  \label{fig:Normalised_Energy_Time_Slope_DEM}
\end{figure}

\begin{figure}[tbhp]
  \centering
  \includegraphics[width=0.9\textwidth]{figs/voro_500}
  \caption{Evolution of packing density with time $E_0 = 152 mgd$ (DEM).}
  \label{fig:voro_500}
\end{figure}

@@

\begin{figure}[tbhp]
  \centering
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/Normalised_KEx_Slope}
    \caption{Evolution of normalised horizontal kinetic energy with time.}
    \label{fig:Normalised_KEx_Slope}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figs/Normalised_KEy_Slope}
    \caption{Evolution of normalised vertical kinetic energy with time.}
    \label{fig:Normalised_KEy_Slope}
  \end{subfigure}
  \caption{Evolution of vertical and horizontal kinetic energy with time (MPM) 
  for a pile subjected to gradient input velocities.}
  \label{fig:Normalised_KEx_KEy_Slope}
\end{figure}

To analyse the second phase for higher input energies, the 
kinetic energy $E'_{kx0}$ at the end of the transient phase is 
considered. This energy is responsible for most of the run-out, hence it is 
expected to 
control the run-out distance and time.~\Cref{fig:Normalised_KExExop_Slope} 
shows the evolution of $E_{kx}$ normalised by $E'_{kx0}$ as a function of time. 
The plots have seemingly the same aspect but they show different decay times. A 
decay time $\tau$ can be defined as the time required for $E_{kx}$ to decline 
by a factor $1/2$.~\Cref{fig:ExEx0_vs_ttau} shows the same data in 
which the time $t'$ elapsed since $t_1$, normalised by $\tau$. Interestingly, 
now all the data nicely collapse on to a single curve. However, this curve can 
not be fitted by simple functional forms such as variants of exponential decay. 
This means that the spreading of the pile is not a self-similar process in 
agreement with the fact that the energy fades away in a finite time $t'_f$.

\clearpage

\begin{figure}[tbhp]
  \centering
  \includegraphics[width=\textwidth]{figs/Normalised_KExExop_Slope}
  \caption[Evolution of the normalised horizontal kinetic energy as function of 
  time since the transient phase.]{Evolution of kinetic energy in the $x$ 
  component of the velocity field normalised by the available kinetic energy at 
  the
  end of the transient phase as a function of time elapsed since the
  same instant (MPM).}
  \label{fig:Normalised_KExExop_Slope}
\end{figure}

\begin{figure}[tbhp]
  \centering
  \includegraphics[width=\textwidth]{figs/EkxKoTTau_Slope}
  \caption[Evolution of the normalised horizontal kinetic energy as function of 
  the normalised time since the transient phase.]{Evolution of kinetic energy in 
  the $x$ component of 
  the velocity field  normalised by the available kinetic energy at the end of 
  the transient as a function of normalised time (MPM).}
  \label{fig:ExEx0_vs_ttau}
\end{figure}

@@

\begin{figure}[tbhp]
  \centering
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/tp_tau_mgd}
    \caption{Power law evolution of $t'_f$ and $\tau$ as a function of kinetic 
    energy $E'_{kx0}$.}
    \label{fig:tp_tau_mgd}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figs/tp_tau}
    \caption{Linear relationship between decay time and run-out time after the 
    transient as a function of the normalised kinetic energy $E_{kx0}$.}
    \label{fig:tp_tau}
  \end{subfigure}
  \caption{Decay time and run-out time as a function of the normalised kinetic 
  energy $E_{kx0}$.}
  \label{fig:tp_Tau}
\end{figure}

@@

CD simulations using different values of coefficient of restitution show no 
difference in the run-out behaviour. At large input energies, the pile remains 
in a dense state so that multiple collisions inside the pile occur at small 
time scales compared to the deformation time. When the restitution coefficients 
are increased, more collisions occur during a longer time interval but the 
overall energy dissipation rate by collisions remains the same. This effect is 
a seminal example of collective effects which erase the influence of local 
parameters at the macroscopic scale.

In contrast to the restitution coefficients, the effect of friction 
coefficient is quite important for the run-out. MPM simulations with 
varying friction coefficient show that, both the run-out distance and the 
decay time decrease as the friction coefficient is increased. This 
effect is much more pronounced at low values of the friction coefficient. 
The run-out time, for example, is reduced by a factor of approximately 4 as 
$\mu_s$ is increased from 0.1 to 0.2 while the change in the run-out and 
duration is less affected with increase in friction coefficient. This 
``saturation effect'' can be observed in a systematic way in simple shear 
tests. The dissipation rate may reach a saturation point where the dilation of 
granular materials and rolling of the grains change in response to increase in 
friction coefficient~\citep{Estrada2008}.

\begin{figure}[tbhp]
  \centering
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figs/runout_fric_slope}
    \caption{Effect of friction on the run-out distance}
    \label{fig:runout_fric_slope}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figs/time_fric_slope}
    \caption{Effect of friction on the duration of run-out.}
    \label{fig:time_fric_slope}
  \end{subfigure}
  \caption{MPM simulations of effect of friction on the run-out behaviour of 
  slopes subjected to horizontal excitation.}
  \label{fig:fric_slope}
\end{figure}

@@

  \item The duration of the flow shows similar behaviour to the run-out, however, 
    a slope subjected to a gradient velocity flows quicker than a slope subjected 
    to a uniform horizontal velocity.

@@

\section*{Acknowledgements}

@@

%%
%% End of file `procs-template.tex'.

>>>>

\title{Transient dynamics of granular slopes: continuum v discrete modelling}

@@

Granular flows have been studied in laboratory experiments in different 
geometries such as tilted piles leading to slope failure and surface
avalanches~\citep{Mutabaruka2015,Legros2002, Iverson1997} or by considering
vertical columns of grains collapsing and spreading under their
ownweight~\citep{Lajeunesse2004, Lajeunesse2005}. The dynamics observed in
theexperiments is often nontrivial in the sense that the final configurations
after the dissipation of the whole kinetic energy can not be readily predicted
by means of simple laws based on the Mohr-Coulomb nature of the material. For
example, in collapsing columns, the runout distance is found to obey a
power-law dependence upon the initial aspect ratio of the column.

@@

In MPM simulations, the material point spacing is adopted to be the same as the 
mean grain diameter in DEM. A mesh size of 0.0125m is adopted with 25 material 
points per cell. The effect of mesh size and the number of material points per 
cell was investigated by~\citep{Soundararajan2015}. The initial configuration
of the slope in MPM is shown in~\cref{fig:mpm_slope_setup}. Frictional boundary
conditions are applied on the left and the bottom boundaries by applying
constraints to the nodal acceleration. Mohr-Coulomb model with no dilation is
used to simulate the continuum behaviour of the granular pile. Periodic shear
tests in DEM reveals a macroscopic friction coefficient of 0.22.

@@

\Cref{fig:mpm_gradvelocity_200j} shows the initial evolution of the 
granular slope subjected to an initial horizontal energy $E_0 = 61$ (in 
dimensionless units) using MPM. As the granular slope is sheared along the 
bottom, the shear propagates to the top leaving a cavity in the vicinity of the 
left wall. This cavity gets partially filled as the granular mass at the top 
collapse behind the flowing mass due to inertia. Similar behaviour is observed 
during the initial stages of the flow evolution using DEM  
(\cref{fig:Gradient_Slope_CD_200J}). Due to inertia, the grains at the top 
of the granular heap roll down to fill the cavity, while the pile continues to 
spread.

\begin{figure}[tbph]
  \centering
  \begin{subfigure}[b]{0.47\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figs/mpm_gradvelocity_200j}
    \caption{MPM}
    \label{fig:mpm_gradvelocity_200j}
  \end{subfigure}
  \begin{subfigure}[b]{0.47\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figs/Gradient_Slope_CD_200J}
    \caption{DEM}
    \label{fig:Gradient_Slope_CD_200J}
  \end{subfigure}
  \caption{Evolution of granular piles subjected to a gradient
    horizontal energy.}
  \label{fig:gradvelocity_200j}
\end{figure}

The flow involves a transient phase with a change in the geometry of the pile 
followed by continuous spreading. The gradient of input energy applied to the 
granular slope mimics a horizontal quake. Despite the creation of a cavity
behind the flowing mass, the granular heap remains in contact with the left
wall irrespective of the input energy.~\Cref{fig:runout_eo_mpm_dem} shows the
normalised run-out distance $(L_f - L_0)/L_0$ and total run-out time $t_f$ as a
function of the input energy $E_0$. Two regimes characterised by a power-law
relation between the run-out distance and time as a function of $E_0$ can be
observed. In the first regime, corresponding to the range of low input energies
$E_0 < 40 \ mgd$, the run-out distance observed varies as $L_f \propto
(E_0)^\alpha$ with $\alpha \simeq 0.206 \pm 0.012$ over nearly one decade.
Overall, the run-out distance predicted by the continuum approach matches the
DEM simulations. At very low energies, DEM simulations show longer run-out
distance due to local fluidisation.

@@

In the second regime, corresponding to the range of high input energies  $E_0 > 
40 \ mgd$, the run-out distance varies as $L_f \propto (E_0)^{\alpha'}$ over 
one decade with $\alpha' \simeq 0.56\pm 0.04$ while the duration increases as 
$t_f \propto (E_0)^{\beta'}$ with $\beta' \simeq 0.33 \pm 0.02$. Hence, in this 
regime the average run-out speed varies as $v_s \propto (E_0)^{0.498 \pm 
0.01}$. This exponent is close to the value $0.5$ in $v_0 \propto (E_0)^{0.5}$, 
and hence, within the confidence interval of the exponents.
In the second regime, both DEM and MPM predict almost the same run-out 
behaviour. However, MPM predicts longer duration with increase in the input 
energy.

@@

The evolution of total kinetic energies $E_k$ with time for different values of 
the input energy $E_{ki}$ based on MPM simulations are shown 
in~\cref{fig:Energy_Time_Slope}. The MPM simulation shows two distinct regimes 
in the normalised kinetic energy plot as a function of normalised time 
in~\cref{fig:Normalised_Energy_Time_Slope_MPM}. However, the DEM simulations 
(\cref{fig:Normalised_Energy_Time_Slope_DEM}) show that the energy evolution 
corresponding to the low energy regime nearly collapse on to a single time 
evolution. This is consistent with the observation of run-out time $t_f$ being 
independent of the input energy. In contrast, MPM simulations predict a power 
law relation between the run-out duration and input energy. However, the 
plots corresponding to the high energy regime (\cref{fig:Energy_Time_Slope}), 
collapse only at the beginning of the run-out i.e. for $t < t_1 \simeq 7.5 \ 
(d/g)^{0.5}$. Although MPM simulations show a longer duration of run-out 
(\cref{fig:Energy_Time_Slope}), the total kinetic energy is completely 
dissipated at $t = 60 \sqrt{d/g}$. DEM simulations predict $t = 80 \sqrt{d/g}$ 
for the kinetic energy to be completely dissipated, which is due to grain 
rearrangement at the free surface. The granular mass densifies as the flow
progresses, after the initial dilation phase for $t = 20\sqrt{d/g}$.

\begin{figure}[tbhp]
  \centering
  \begin{subfigure}[b]{0.4\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figs/Normalised_Energy_Time_Slope_MPM}
    \caption{MPM}
    \label{fig:Normalised_Energy_Time_Slope_MPM}
  \end{subfigure}
  \begin{subfigure}[b]{0.4\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figs/Normalised_Energy_Time_Slope_DEM}
    \caption{DEM}
    \label{fig:Normalised_Energy_Time_Slope_DEM}
  \end{subfigure}
  \caption{Evolution of normalised kinetic energy with normalised time for a
  pile subjected to gradient input velocities.}
  \label{fig:Energy_Time_Slope}
\end{figure}

@@

\begin{figure}[tbhp]
  \centering
  \begin{subfigure}[b]{0.4\textwidth}
    \includegraphics[width=\textwidth]{figs/Normalised_KEx_Slope}
    \caption{Evolution of normalised horizontal kinetic energy with time.}
    \label{fig:Normalised_KEx_Slope}
  \end{subfigure}
  \begin{subfigure}[b]{0.4\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figs/Normalised_KEy_Slope}
    \caption{Evolution of normalised vertical kinetic energy with time.}
    \label{fig:Normalised_KEy_Slope}
  \end{subfigure}
  \caption{Evolution of vertical and horizontal kinetic energy with time (MPM) 
  for a pile subjected to gradient input velocities.}
  \label{fig:Normalised_KEx_KEy_Slope}
\end{figure}

To analyse the second phase for higher input energies, the kinetic energy
$E'_{kx0}$ at the end of the transient phase is considered. This energy is
responsible for most of the run-out, hence it is expected to control the
run-out distance and time. A decay time $\tau$ can be defined as the time
required for $E_{kx}$ to decline by afactor $1/2$.~\Cref{fig:ExEx0_vs_ttau}
shows the same data in which the time $t'$ elapsed since $t_1$, normalised by
$\tau$. Interestingly, now all the data nicely collapse on to a single curve.
However, this curve can not be fitted by simple functional forms such as
variants of exponential decay. This means that the spreading of the pile is not
a self-similar process in agreement with the fact that the energy fades away in
a finite time $t'_f$.

\begin{figure}[tbhp]
  \centering
  \includegraphics[width=0.4\textwidth]{figs/EkxKoTTau_Slope}
  \caption[Evolution of the normalised horizontal kinetic energy as function of 
  the normalised time since the transient phase.]{Evolution of kinetic energy in 
  the $x$ component of the velocity field  normalised by the available kinetic
  energy at the end of the transient as a function of normalised time (MPM).}
  \label{fig:ExEx0_vs_ttau}
\end{figure}

@@

\begin{figure}[tbhp]
  \centering
  \begin{subfigure}[b]{0.4\textwidth}
    \includegraphics[width=\textwidth]{figs/tp_tau_mgd}
    \caption{}
    \label{fig:tp_tau_mgd}
  \end{subfigure}
  \begin{subfigure}[b]{0.4\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figs/tp_tau}
    \caption{}
    \label{fig:tp_tau}
  \end{subfigure}
  \caption{Decay time and run-out time as a function of the normalised kinetic 
  energy $E_{kx0}$: (a) Power law evolution of $t'_f$ and $\tau$ as a function
  of kinetic energy $E'_{kx0}$ and (b) Linear relationship between decay time
  and run-out time after the transient as a function of the normalised kinetic
  energy $E_{kx0}$.}
  \label{fig:tp_Tau}
\end{figure}

@@

The effect of friction coefficient is quite important for the run-out. MPM
simulations with varying friction coefficient show that, both the run-out
distance and the decay time decrease as the friction coefficient is increased.
This effect is much more pronounced at low values of the friction coefficient.
The run-out time, for example, is reduced by a factor of approximately 4 as
$\mu_s$ is increased from 0.1 to 0.2 while the change in the run-out and
duration is less affected with increase in friction coefficient. This
``saturation effect'' can be observed in a systematic way in simple shear
tests. The dissipation rate may reach a saturation point where the dilation of
granular materials and rolling of the grains change in response to increase in
friction coefficient~\citep{Estrada2008}.

\begin{figure}[tbhp]
  \centering
  \begin{subfigure}[b]{0.4\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figs/runout_fric_slope}
    \caption{run-out distance}
    \label{fig:runout_fric_slope}
  \end{subfigure}
  \begin{subfigure}[b]{0.4\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figs/time_fric_slope}
    \caption{duration of run-out}
    \label{fig:time_fric_slope}
  \end{subfigure}
  \caption{MPM simulations of effect of friction on the run-out behaviour of 
  slopes subjected to horizontal excitation.}
  \label{fig:fric_slope}
\end{figure}

@@

@@

\section*{Acknowledgements}
The authors would like to thank Professor F. Radjai, LMGC, Montpellier, France for stimulating discussions regarding this work. The author would like to thank the Cambridge Commonwealth and Overseas Trust for the financial support to the first author to pursue this research.

@@

%% End of file `procs-template.tex'.
